Predicting Survival Probability of NASA Aircraft Engines

### 2025-04-13 index.qmd
Slides Update 4/13

---
title: "Predicting Survival Probability of NASA Aircraft Engines"
subtitle: "Using Cox Proportional Hazards Model"
author: "Jayme Reed & Brad Paton (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    theme: spacelab
    toc: true
    toc-depth: 3
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

@@

Depending on the study, there will be covariates that do not impact the
probability of survival. Statistical methods such as the Wald Test can
be used to determine which ones to use in the final CPH model.
Generating a univariate CPH model for each covariate and using the Wald
test with a significance level of $\alpha = 0.05$ will allow for
determination of if a covariate is significant. Any covariates that are
determined to be not significant to the survival probability can be
removed from the final CPH model.

>>>>

---
title: "Predicting Survival Probability of NASA Aircraft Engines"
subtitle: "Using Cox Proportional Hazards Model"
author: "Jayme Reed & Brad Paton (Advisor: Dr. Cohen)"
date: last-modified
date-format: long
format:
  html:
    theme: spacelab
    toc: true
    toc-depth: 3
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

@@

Depending on the study, there will be covariates that do not impact the
probability of survival. Statistical methods such as the Wald Test can
be used to determine which ones are statistically significant
individually and may have an impact in a multivariate model. Generating
a univariate CPH model for each covariate and using the Wald test with a
significance level of $\alpha = 0.05$ will allow for determination of if
a covariate is significant. Additional tests for models will involve
using a step-wise regression and comparing the results based on the
assumptions that are met and the concordance index. Any covariates that
are determined to be not significant to the survival probability can be
removed from the final CPH model.

### 2025-04-13 slides.qmd
Slides Update 4/13

---
title: "Predicting Survivability of NASA Aircraft Engines"
subtitle: "Using Cox Proportional Hazards Model"
author: "Jayme Reed & Brad Paton (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs:
    theme: moon
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

## Data Exploration and Visualization

## Data Exploration and Visualization {.smaller}

## Modeling and Results

>>>>

---
title: "Predicting Survival Probability of NASA Aircraft Engines"
subtitle: "Using Cox Proportional Hazards Model"
author: "Jayme Reed & Brad Paton (Advisor: Dr. Cohen)"
date: last-modified
date-format: long
format:
  revealjs:
    theme: moon
    smaller: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
  echo: false
editor: 
  markdown: 
    wrap: 72
---

```{r, echo = FALSE, warning=FALSE, include=FALSE}
#Load packages
library(tidyverse)
library(gridExtra)
library(grid)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)
library(sjPlot)
library(broom)
```

```{r, echo = FALSE}
#generation of the model for sample plots
example_surv_object <- Surv(time = lung$time, event = lung$status)

example_coxmodel <- coxph(example_surv_object ~ meal.cal, data = lung)
```

## Cox Proportional Hazards (CPH) Model

What is it?

-   A statistical regression method specializing in modeling
    time-to-event predictions with survival data [@Abeysekera2009]
    -   Survival data has a value for time and an indicator column for
        an event
-   Is a method that can deal with censored data
    -   Censored data is when the information about an individual in a
        study is only known for a certain period of time [@klein2005]
-   Primarily used in the health field but has applications in
    predicting bank failure, the survival probability of machines, and
    insurance likelihood payouts

## Limitations

-   The model assumes that as time goes on, the survival probability
    will approach zero with no survivors [@asghar2024]

-   The proportional hazards assumption can limit the ability to
    correctly predict the effect of a variable [@jiang2024]

-   The covariate selection can become biased and may not accurately
    represent the true data [@wang2025],[@zhang2025]

-   The model cannot provide a specific value for when the event will
    happen, only the probability of when the event might happen

## Mathematical Formulas

-   Survival Function:

    $S(x) = \int_{x}^{\infty} f(x)dx$

-   Hazard Function:

    $h(x) = \lim\limits_{\Delta x \rightarrow 0} \frac{P[x \leq X < x + \Delta x|X \geq x]}{\Delta x}$

-   CPH Model Hazard Function:

    $h(t|\mathbf Z) = h_0(t)\text{exp}(\sum\limits_{k=1}^{p} \beta_kZ_k)$

-   Proportional Hazards Ratio:

    $\frac{h(t|\mathbf Z)}{h(t|\mathbf Z*)} = \text{exp}[\sum\limits_{k=1}^{p} \beta_k(Z_k - Z_k^*)]$

-   Cumulative Hazard Function:

    $H(x) = \int_0^x h(u) du$

## Assumptions

There are four assumptions for CPH:

::: incremental
-   Independence assumption

    -   Assumes that the survival times of observed subjects are
        independent of each other [@nahhas2025]

-   Non-informative Censoring Assumption

    -   Assumes that censoring is non-informative [@nahhas2025]
:::

## Assumptions, Cont.

-   Linearity Assumption

    ::: incremental
    -   Assumes the relationship between covariates and the outcome is a
        linear relationship

    -   This is tested using the Martingale residuals for each covariate
        with the equation Martingale Residuals = Observed Events -
        Expected Results [@nahhas2025]

        -   The residuals are examined visually

        -   If linear and appear to have a slop of zero, the assumption
            is not violated [@Amini2015]
    :::

```{r,echo=FALSE}
example_martingale <- resid(example_coxmodel, type = "martingale")

plot(example_martingale, 
     main = "Martingale Residuals: Example Covariate",
     ylab = "Martingale Residuals for Example Covariate",
     xlab = "Value Range for Example Covariate")
abline(h = 0, col = "red")

```

## Assumptions, Cont.

-   Proportional Hazards Assumption

    ::: incremental
    -   The ratio of hazard rates for any two subjects must be constant
        at all times [@Bustan2018]

    -   To test this assumption, the below hypothesis test is used:

        $$
        \begin{align*}
        H_0&:\; \text{the log hazard ratio for each covariate remains constant over time} \\
        H_1&: \; \text{the log hazard ratio for each covariate does not remain constant over time}
        \end{align*}
        $$

        -   For this test, the goal is to *not* reject $H_0$

    -   The scale Schoenfeld partial residuals are used to find the
        p-value for this test

        -   The Schoenfeld residuals are the difference between the
            value of the covariate and the expected value of the
            covariate at the time of failure [@klein2005]

        -   There will be multiple p-values generated: global and
            individual

            -   If the global p-value is $< 0.05$, then the model has at
                least one covariate that violates the proportional
                hazards assumption and the individual p-values must be
                examined

            -   If the global p-value is $> 0.05$, then all covariates
                meet the proportional hazard assumption [@nahhas2025]

    Example of Schoenfeld plot and results

    ```{r, echo = FALSE}
    example_phtest <- cox.zph(example_coxmodel)

    plot(example_phtest[1],
         ylab = "Covariate Schoenfeld Residual",
         xlab = "Time")

    multi <- coxph(example_surv_object ~ meal.cal + age, data = lung)
    ph_tibble <- cox.zph(multi) %>%
      pluck("table") %>%
      as.data.frame() %>%
      tibble::rownames_to_column("variable") %>%
      as_tibble()
    kable(ph_tibble, format = "markdown")

    ```
    :::

## Model Visualization - Kaplan Meier

## Model Visualization - Forest Plot

## Evaluation

## Prediction

## Data Structure, Exploration, and Visualization

## Creating CPH Model

## Checking Assumptions

## Adjusting the Model

## Model Results
